Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2020-11-25 sources/bmc_template_submission/bmc_article.tex
italics

More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of `r round(236023/240708, digits=4)*100`% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.
Assembling a DNA alignment from such a massive database can be done "by hand",
but it is a largely time consuming and mostly non-reproducible approach.
Computational pipelines that mine DNA databases fast, efficiently, and reproducibly,
have been applied to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper}).
However, fine-grained curated markers and alignments can improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

@@

A user is interested in phylogenetic relationships within the genus *Ilex*. Commonly
known as "hollies", the genus encompasses between 400-700 living species, and is
the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering
plants.

@@

The most recent *Ilex* tree \cite{yao2020phylogeny} is available in
OpenTree's Phylesystem \href{https://tree.opentreeoflife.org/curator/study/view/ot_1984}{study ot\_1984}
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the \cite{yao2020phylogeny} tree is the best sampled phylogeny yet available for
the hollies.

We ran Physcraper on a laptop Linux computer to update an internal transcribed
spacer DNA region (ITS) alignment from \cite{gottlieb2005molecular}, using a local
GenBank database.  BLAST and RAxML analyses ran for 19hrs 45min, with bootstrap
analyses taking an additional 13hrs.
<!-- **MTH: you probably need some details about the hardware, given the fact
hat you are discussing running times** -->
The updated \cite{gottlieb2005molecular} tree (Fig. \@ref(fig:results)) displays
all 41 distinct
taxa from the original study plus 231 new tips, contributing phylogenetic data to
84 additional *Ilex* taxa. The best RaxML tree is 99% resolved, with 25% of nodes
with bootstrap support < 0.1 and 48% nodes with bootstrap support > 0.75.
A large portion of internal branches are negligibly small, with 30 branches
< 0.00001 substitution rate units, from which only 9 have a bootstrap support > 0.75
(Figure 2).
For comparison, \cite{yao2020phylogeny} also contains all 41 distinct taxa from
the original
\cite{gottlieb2005molecular} study,
and contributes phylogenetic data to 134 additional *Ilex* taxa, from which
`r 135-68` are also in updated \cite{gottlieb2005molecular}. While \cite{yao2020phylogeny} also used
ITS as a marker, their GenBank data is not released yet, so Physcraper was unable
to incorporate 68 additional taxa into the analysis. However, Physcraper was able
to incorporate 18 taxa that were not in \cite{yao2020phylogeny}.

@@

Phylogenetic pipelines designed to make evolutionary sense of the vast amount of
public molecular data (e.g., Phylota \cite{sanderson2008phylota},
PHLAWD \cite{smith2009mega},
SUPERSMART \cite{antonelli2017toward}) focus on generating full phylogenies *de novo*,
i.e., inferring phylogenetic relationships from a newly generated homology hypothesis,
as opposed to e.g., supertrees, that are generated by summarizing previous
phylogenetic estimates.
While Physcraper does not generate phylogenies *de novo* in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the importance of
open science in facilitating phylogenetic placement of public molecular data and
accelerating enrichment and updating of phylogenetic relationships in any region
of the tree of life.
The PUMPER pipeline \cite{izquierdo2014pumper} also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies.
Unfortunately, installation was unsuccessful following instructions from the author,
and a comparison analysis between Physcraper and PUMPER is unfeasible at present.
%% **MTH: add some more details about what aspect of the PUMPER install failed**

>>>>

More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of 98.05\% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.
Assembling a DNA alignment from such a massive database can be done "by hand",
but it is a largely time consuming and mostly non-reproducible approach.
Computational pipelines that mine DNA databases fast, efficiently, and reproducibly,
have been applied to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper}).
However, fine-grained curated markers and alignments can improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

@@

A user is interested in phylogenetic relationships within the genus \textit{Ilex}. Commonly
known as "hollies", the genus encompasses between 400-700 living species, and is
the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering
plants.

@@

The most recent \textit{Ilex} tree \cite{yao2020phylogeny} is available in
OpenTree's Phylesystem \href{https://tree.opentreeoflife.org/curator/study/view/ot_1984}{study ot\_1984}
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the \cite{yao2020phylogeny} tree is the best sampled phylogeny yet available for
the hollies.

We ran Physcraper on a laptop Linux computer to update an internal transcribed
spacer DNA region (ITS) alignment from \cite{gottlieb2005molecular}, using a local
GenBank database.  BLAST and RAxML analyses ran for 19hrs 45min, with bootstrap
analyses taking an additional 13hrs.
%% <!-- MTH: you probably need some details about the hardware, given the fact
hat you are discussing running times -->
The updated \cite{gottlieb2005molecular} tree (Figure 2) displays
all 41 distinct
taxa from the original study plus 231 new tips, contributing phylogenetic data to
84 additional \textit{Ilex} taxa. The best RaxML tree is 99% resolved, with 25% of nodes
with bootstrap support < 0.1 and 48% nodes with bootstrap support > 0.75.
A large portion of internal branches are negligibly small, with 30 branches
< 0.00001 substitution rate units, from which only 9 have a bootstrap support > 0.75
(Figure 2).
For comparison, \cite{yao2020phylogeny} also contains all 41 distinct taxa from
the original
\cite{gottlieb2005molecular} study,
and contributes phylogenetic data to 134 additional \textit{Ilex} taxa, from which
`r 135-68` are also in updated \cite{gottlieb2005molecular}. While \cite{yao2020phylogeny} also used
ITS as a marker, their GenBank data is not released yet, so Physcraper was unable
to incorporate 68 additional taxa into the analysis. However, Physcraper was able
to incorporate 18 taxa that were not in \cite{yao2020phylogeny}.

@@

Phylogenetic pipelines designed to make evolutionary sense of the vast amount of
public molecular data (e.g., Phylota \cite{sanderson2008phylota},
PHLAWD \cite{smith2009mega},
SUPERSMART \cite{antonelli2017toward}) focus on generating full phylogenies \textit{de novo},
i.e., inferring phylogenetic relationships from a newly generated homology hypothesis,
as opposed to e.g., supertrees, that are generated by summarizing previous
phylogenetic estimates.
While Physcraper does not generate phylogenies \textit{de novo} in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the importance of
open science in facilitating phylogenetic placement of public molecular data and
accelerating enrichment and updating of phylogenetic relationships in any region
of the tree of life.
The PUMPER pipeline \cite{izquierdo2014pumper} also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies.
Unfortunately, installation was unsuccessful following instructions from the author,
and a comparison analysis between Physcraper and PUMPER is unfeasible at present.
%% **MTH: add some more details about what aspect of the PUMPER install failed**
